Kindly Bent to Free Us

### 2019-06-17 kindly/motivation.tex
Write the sudoku section.

While purely functional data-structure present numerous advantages in term
of correctness and ease of use, they sometimes offer worse performances
than their imperative counterpart. On the other hand, reasoning about imperative
data-structure in the context of sharing is notoriously difficult.
%
In this section, we show how linearity and borrowing allows us to define
a safe API for hybrid arrays that support both immutable and mutable features,
and how it can be used to implement a backtracking sudoku solver.

@@

  val set : &('a t) * int * 'a -> 'a t
  val set_immut : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Signature}
    \label{sig:hybarray}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
module Cow : HybArray = struct
  include Array

  let set (a, i, x) =
    let f k =
      if k = i then x else Array.get (a,k)
    in
    create (length a) f

@@

\begin{figure}
  \centering
\begin{lstlisting}[multicols=2]
type board = IntSet.t Mat.t

let propagate_line i0 j0 g n =
  for j = j0+1 to size - 1 do
    let cell = Mat.get &g (i0 , j) in
    let cell' = IntSet.remove n cell in
    Mat.set_mut &!g (i0, j) cell'
  done

let propagate i j g n =
  propagate_line i j &!g n;
  propagate_column i j &!g n;
  propagate_square i j &!g n

let rec solve i j g =
  match is_solved &g with
  | true -> Mat.print &g
  | false ->
    let (new_i, new_j) = next_pos (i,j) in
    let try_solution n =
      let new_g = 
        Mat.set &g (i,j) (singleton n) in
      propagate i j &!new_g n;
      if is_valid &new_g when
        solve new_i new_j new_g
    in
    let cell = Mat.get &g (i,j) in
    IntSet.iter try_solution cell
\end{lstlisting}
  \caption{Chosen excerpt from the sudoku solver}
  \label{ex:cow}
\end{figure}

>>>>

In the last few years, new persistent data-structures
have been introduced which allow transient mutations, but discourage
non-linear uses either through degraded performances
\cite{DBLP:conf/ml/ConchonF07} or
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Trie has been used with similar
API in numerous non-pure functional languages (OCaml, Clojure, \dots).
Affine types allow to formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an immutable API.
%
In this section, we show how linearity and borrowing allows us to define
a safe API for persistent arrays that support both immutable and mutable features,
and how it can be used to implement a backtracking sudoku solver.

Our sudoku solver is implemented by maintaining a grid that represent the state of the game. In order to make backtracking easier, we want to modify the grid immutably. However, while some changes correspond to backtracking points,
some do not.
In order to improve performance, we want to implement modifications
that do not correspond to backtracking points as mutations.
Thanks to affine types and borrows, we can provide an API that guarantee correct
uses of mutable and immutable modifications to the board.
This API is represented by the signature \lstinline/HybArray/ in \cref{sig:hybarray}.
It only differs with the \lstinline/Array/ module on two points.
The mutation function is now called \lstinline/set_mut/, with
a type \lstinline/&!('a t) * int * 'a -> unit/ and
the \lstinline/set/ operation
is now of type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an \emph{immutable} borrow, which allows it to read the array, and
return a fresh hybrid array.
We showcase a very simple implementation of \lstinline/HybArray/ in
\cref{ex:cow} where hybrid arrays are represented
as regular arrays with copy-on-write for immutable
modifications.
A more elaborate implementation could use persistent arrays
\cite{DBLP:conf/ml/ConchonF07}.

@@

  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Signature}
    \label{sig:hybarray}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
module Cow : HybArray = struct
  include Array

  let set (a, i, x) =
    let f k =
      if k = i then x else get (a,k)
    in
    create (length a) f

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.

While we presented our API on simple copy-on-write arrays, this API
could easily be adapted to any persistent data-structure with
transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
and persistent HAMT (used notably as Clojure's default associative maps
\cite{bagwell2001ideal,clojurehamt}) while providing a good performance profile
and a convenient programming style.

\begin{figure}
  \centering
  \begin{minipage}[t]{0.45\linewidth}
    \lstinputlisting[linerange=1-13]{code/sukodu.affe}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.54\linewidth}
    \lstinputlisting[linerange=15-28,firstnumber=last]{code/sukodu.affe}
  \end{minipage}
  \caption{Chosen excerpt from the sudoku solver}
  \label{ex:sudoku}
\end{figure}
